---
title: "PSTAT 126 Final Project"
author: "Amy Nguyen (A244C16)"
output: pdf_document
date: "2026-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
# Part 1: Data Description and Descriptive Statistics

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(corrplot)
```

## 1: Select a Random Sample
```{r}
set.seed(1432)
data(diamonds)
diamond_sample <- diamonds %>% sample_n(5000)
diamond_df <- data.frame(diamond_sample)
```
\newpage
## 2: Describe the Variables
### Summary of Data
```{r}
str(diamond_df)
summary(diamond_df)
```
\newpage
### Histograms of Continuous Variables
```{r, fig.width=6, fig.height=4}
hist(diamonds$carat, main='Histogram of Diamond Weight (Carat)', xlab='Weight (Carat)')
```
\
Carat: The distribution of **`carat`** is right-skewed, with most diamonds being less than 1 carat. \
```{r, fig.width=6, fig.height=4}
hist(diamonds$depth, main='Histogram of Diamond Depth Percentage', xlab='Depth Percentage')
```
\
Depth Percentage: The distribution of **`depth`** is roughly symmetric, with a central tendency around 61%.
\newpage
```{r, fig.width=6, fig.height=4}
hist(diamonds$table, main='Histogram of Diamond Table Percentage', xlab='Table Percentage')
```
\
Table Percentage: The distribution of **`table`** is roughly symmetric, with a central tendency around 58%. \
```{r, fig.width=6, fig.height=4}
hist(diamonds$price, main='Histogram of Diamond Price ($)', xlab='Price ($)')
```
\
Price: The distribution of **`price`** is extremely right-skewed, with most diamonds being around 1000-2000 US dollars. 
\newpage
```{r, fig.width=6, fig.height=4}
hist(diamonds$x, main='Histogram of Diamond Length x (mm)', xlab='Length x (mm)')
```
\
Length: The distribution of **`x`** is slightly right-skewed, with most diamonds being around 4 mm long. \
```{r, fig.width=6, fig.height=4}
hist(diamonds$y, main='Histogram of Diamond Width y (mm)', xlab='Width y (mm)', breaks=30)
```
\
Width: The distribution of **`y`** is slightly right-skewed, with most diamonds being around 4-5 mm wide.
\newpage
```{r, fig.width=6, fig.height=4}
hist(diamonds$z, main='Histogram of Diamond Depth z (mm)', xlab='Depth z (mm)', breaks=30)
```
\
Depth: The distribution of **`z`** is slightly right-skewed, with most diamonds being around 3-4 mm deep. \
\newpage

### Barplots of Categorical Variables
```{r, fig.width=5, fig.height=3}
ggplot(data=diamond_df, aes(x=cut)) +
  geom_bar() +
  labs(title='Barplot of Diamond Cut', x='Cut', y='Count') +
  theme(plot.title = element_text(hjust = 0.5))
```
\
Cut: **`Ideal`** seems to be the most common cut of diamond, while **`Fair`** is the most rare. \
```{r, fig.width=5, fig.height=3}
ggplot(data=diamond_df, aes(x=color)) +
  geom_bar() +
  labs(title='Barplot of Diamond Color', x='Color', y='Count') +
  theme(plot.title = element_text(hjust = 0.5))
```
\
Color: The distribution of **`color`** is roughly uniform, where **`G`** seems to be the most common color, which makes sense as it represents an average diamond color.
\newpage
```{r, fig.width=5, fig.height=3}
ggplot(data=diamond_df, aes(x=clarity)) +
  geom_bar() +
  labs(title='Barplot of Diamond Clarity', x='Clarity', y='Count') +
  theme(plot.title = element_text(hjust = 0.5))
```
\
Clarity: **`SI1`** (Slightly Included 1) and **`VS2`** (Very Slightly Included 2) are the most common grades of diamonds.
\newpage

## 3: Select 3 Quantitative and 2 Categorical Variables
### Quantitive Variables: Carat, Depth, Table
### Categorical Variables: Cut, Clarity
```{r}
vars <- diamond_df %>%
  mutate(cut_int = as.integer(cut), clarity_int = as.integer(clarity)) %>%
  select(carat, depth, table, cut_int, clarity_int)
corr_matrix <- cor(vars)
print(corr_matrix)
```
\
```{r, fig.width=4, fig.height=4}
corrplot(corr_matrix, method='color', type='upper', tl.col='black', addCoef.col = "green",
         number.cex = 0.75,
         title = "Correlation Matrix of Selected Variables",
         mar = c(0, 0, 1, 0))
```
\
The heatmap shows that **`table`** and **`cut`** have the strongest negative correlation (-0.43), with **`carat`** and **`clarity`** also having a moderate negative correlation (-0.36). There are weak positive correlations between **`cut`** and **`clarity`** (0.20), as well as **`carat`** and **`table`** (0.17). There are weak negative correlations between **`depth`** and **`table`** (-0.28), **`depth`** and **`cut`** (-0.24), and **`table`** and **`clarity`** (-0.16). The rest of the variables have very weak correlations with each other.
\newpage

## 4: Run Multiple Linear Regression Model
```{r}
mlr <- lm(price ~ carat + depth + table + cut + clarity, data=diamond_df)
summary(mlr)
```
\newpage
## 5: Observations
It was interesting how the selected variables had mostly low correlations between each other. In particular, I was surprised that **`carat`** and **`depth`** were so lowly correlated (0.03), because I had originally thought that the weight of the diamond would have some dependency on the depth percentage. I am not surprised, though, that **`carat`** and **`clarity`** have a moderate negative correlation (-0.36) because the bigger the diamond, the more flaws (more inclusions, lower clarity) that it will have. 